Extracting Angular Observables without a Likelihood\\and Applications to Rare Decays

### 2015-05-07 draft.tex
added my second affiliation
added acknowlegment
[draft] Add references on old work
* update acknowledgment
* minor updates of formulae
* add questions for less minor changes

%% Draft Macros %%
%\newcommand{\todo}[1]{{\color{red}\bf ToDo: #1}}
%\newcommand{\danny}[1]{{\color{purple}#1}}
%\newcommand{\fred}[1]{{\color{brown!85!black}#1}}
%\newcommand{\marcin}[1]{{\color{cyan}#1}}
%\newcommand{\citeneeded}{{\color{red}\bf [cite needed]}}
%\newcommand{\add}[1]{\textcolor{green!85!black}{#1}}
%\newcommand{\del}[1]{\textcolor{red!85!black}{#1}}

@@

\preprint{SI-HEP-2014-17}
\title{Extracting Angular Observables without a Likelihood\\and Applications to Rare Decays}
\author{Frederik Beaujean}
\email{frederik.beaujean@lmu.de}
\affiliation{C2PAP, Universe Cluster, Ludwig-Maximilians-Universit\"at M\"unchen, Garching, Germany}
\author{Marcin Chrz\k{a}szcz}
\email{mchrzasz@cern.ch}
\author{Nicola Serra}
\email{nicola.serra@cern.ch}
\affiliation{Physik-Institut, Universit\"at Z\"urich, Z\"urich, Switzerland}
\author{Danny van Dyk}
\email{vandyk@tp1.physik.uni-siegen.de}
\affiliation{Theoretische Physik 1, Naturwissenschaftlich-Technische Fakult\"at,
Universit\"at Siegen, Siegen, Germany}

@@

Our method is an extension of the classical method of moments with
orthogonal functions~\cite[sec. 8.2]{James:2006zz}. The only difference
is that conventionally the angular functions are assumed
\emph{self-dual}, $\dual{f}_i = f_i$. However, it suffices that the
system of angular functions $\{f_i(\thvec)\}$ \emph{can} be transformed into
an orthonormal basis. We find it convenient to work
in the basis of Legendre polynomials that are not self-dual.
Our approach covers the self-dual case, provided that one
replace $\dual{f}_i \to f_i$ appropriately.
Using the ansatz
\begin{equation}
  \label{eq:dual-ansatz}
  \dual{f}_i = \sum_{j} a_{ij} f_j \,,
\end{equation}
the dual basis needs to be worked out case by case through solving the
linear system of equations (\ref{eq:def-ortho-rel}). For a selection
of hadron decays with a $b$ quark in the initial state and two leptons
in the final state, we list the dual bases in a series of appendices
\ref{app:btokll} through \ref{app:btokstarll}. Note that a
similar analysis was done in~\cite{Dighe:1998vk} for the decays $B \to
J/\psi \phi$ and $B \to J/\psi K^{*}$.

@@

The orthonormality relations \refeq{def-ortho-rel} imply that a single angular observable $S_i$
can be projected out of the full PDF $P$ as
\begin{equation}
    \label{eq:det-Pi-analytical}
    S_i(\nuvec) = \int_{\Omega} \rmdx{\vec\theta}  P(\nuvec, \thvec) \dual{f}_i(\thvec)\,.
\end{equation}
where $\lbrace \dual{f}_i \rbrace$ denotes a \emph{dual basis} of angular
functions, and $\Omega$ represents the entire angular phase space. In general,
$\lbrace \dual{f}_i \rbrace$ may differ from $\lbrace f_i \rbrace$. This is the
case for our selection of applications in appendices \ref{app:btokll} through
\ref{app:btokstarll}.\\

It is sensible to refer to the angular observable $S_i$ as the
\emph{$f_i$-moment} of the PDF $P$.  We emphasize that a relation of
type \refeq{det-Pi-analytical} holds for any combination of a density
written as in \refeq{def-P} and an orthonormal basis of angular
functions $\lbrace f_i \rbrace$; \ie, there is no unique
basis of angular functions. For the proof we refer to ref. \cite{Dighe:1998vk}.\\

@@

In what follows, we propose a systematic method to unfold all effects
of $\Ekernel$ through MC simulations and the method of moments,
using that $\Ekernel$ can be expanded --- at
least formally --- in Legendre polymials.  For illustration, we
proceed with the explicit example of a uniangular\footnote{%
  The generalization of this section to multiangular PDFs is
  straighforward. It can be achieved by promoting $x$ to a vector,
  promoting the Legendre polynomials to products of independent
  polynomials or spherical harmonics, and promoting the indices
  $i,j,k,m$ to multi-indices.  } PDF with $x = \cos\theta$. Let us
define the PDF in terms of the Legendre polymials (i.e., $f_k(x)
\equiv p_k(x)$) and angular observables $\vec{S}$ as
\begin{equation}
    P\subt(\xt) \equiv P\subt(\xt | \vec{S}) = \sum_k S_k f_k(\xt),
  \end{equation}
where $k = 0, 1, \dots$ denotes an angular-momentum-like index associated with the observables.
Normalization of $P\subt$ is equivalent to choosing $S_0 = 1/2$. Requiring $P\subt(\xt) \ge 0 \,\forall \xt$
implies $|S_k| \leq 1/2$ for $k > 0$. More stringent relations between the
$S_k$ might hold, but are of no concern here. For later use, we define
\begin{equation}
  \label{eq:def-Sk}
  S^{(m)}_k \equiv
  \begin{cases}
    1/2 \delta_{k,0}, & m=0\\
    1/2 (\delta_{k,0} + \delta_{k,m}), & m>0 \,,
  \end{cases}
\end{equation}
and note that
\begin{equation}
    P^{(m)}(\xt) \equiv P(\xt | \vec{S}^{(m)})
\end{equation}
is a valid PDF. The dual basis of angular functions follows then from the normalization
and orthogonality of the Legendre polynomials, and one therefore has
\begin{gather}
    \tilde{f}_k(x) = \frac{2 k + 1}{2} f_k(x)\,,\\
    \int_{-1}^{+1} \rmdx{x} \tilde{f}_k(x) f_l(x) = \delta_{k,l}\,.
\end{gather}
We now define the \emph{simulated raw} moments $\vec{Q}^{(m)}$ as
\begin{equation}
    \label{eq:analytic-raw-moments}
    Q_i^{(m)} \equiv \iint \rmdx{\xt} \rmdx{\xd} \tilde{f}_i(\xd) P^{(m)}(\xt) \Ekernel,
\end{equation}
which are instrumental to our recipe. Monte Carlo estimators of these moments can be constructed from specifically crafted detector events $x_\text{d}^{(n,m)}$, $n = 1, \dots, N_\text{d}^{(m)}$, where
\begin{equation}
    x_\text{d}^{(n,m)} \sim P_E^{(m)}(\xd) \equiv \frac{1}{R^{(m)}} \int \rmdx{\xt} P^{(m)}(\xt) \Ekernel\,.
\end{equation}
In words, for each $m$ it is required to generate events from
a toy physical distribution $P^{(m)}$, for which $S_0 = S_m=1/2$ and all other
observables are set to zero. Next, propagate these events through
a detector simulation. The normalization $R^{(m)}$ is chosen such
that $\int P_E^{(m)}(\xd) \rmdx{\xd} = 1$.  We emphasize that
$R^{(m)}$ can be estimated as $\est{R}^{(m)} = N_\text{d}^{(m)} /
N_\text{t}$, where $N_\text{t}$ corresponds to the number of
simulated true events.  The estimators then read
\begin{equation}
    \est{Q}_i^{(m)} \equiv \est{R}^{(m)} \frac{1}{N_\text{d}} \sum_n^{N_\text{d}} \tilde{f}_i(x_\text{d}^{(n,m)})
\end{equation}

@@

@@

We thank Ulrik Egede for insightful discussions and helpful comments
on the manuscript. We are grateful to Gudrun Hiller, Martin Jung, and
Konstantinos Petridis for helpful discussions, as well as Claus Grupen
for helpful advice in searching the literature.  The work of D.v.D has
been supported by the Bundesministerium f\"ur Bildung und Forschung
(BMBF). F.B. is grateful for the support by the DFG Cluster of
Excellence "Origin and Structure of the Universe". Part of the
numerical simulations have been carried out on the computing
facilities of the Computational Center for Particle and Astrophysics
(C2PAP).

>>>>

%% Draft Macros %%
%\newcommand{\todo}[1]{{\color{red}\bf ToDo: #1}}
%\newcommand{\danny}[1]{{\color{purple}#1}}
\newcommand{\fred}[1]{{\color{brown!85!black}#1}}
%\newcommand{\marcin}[1]{{\color{cyan}#1}}
%\newcommand{\citeneeded}{{\color{red}\bf [cite needed]}}
%\newcommand{\add}[1]{\textcolor{green!85!black}{#1}}
%\newcommand{\del}[1]{\textcolor{red!85!black}{#1}}

@@

\preprint{SI-HEP-2014-17}
\title{Extracting Angular Observables without a Likelihood\\and Applications to Rare Decays}
\author{Frederik Beaujean}
\email{frederik.beaujean@lmu.de}
\affiliation{C2PAP, Universe Cluster, Ludwig-Maximilians-Universit\"at M\"unchen, Garching, Germany}
\author{Marcin Chrz\k{a}szcz}
\email{mchrzasz@cern.ch}
\affiliation{Physik-Institut, Universit\"at Z\"urich, Z\"urich, Switzerland \\ and \\ H.Niewodniczanski Institute of Nuclear Physics, Polish Academy of Sciences, Krakow, Poland}
\author{Nicola Serra}
\email{nicola.serra@cern.ch}
\affiliation{Physik-Institut, Universit\"at Z\"urich, Z\"urich, Switzerland}
\author{Danny van Dyk}
\email{vandyk@tp1.physik.uni-siegen.de}
\affiliation{Theoretische Physik 1, Naturwissenschaftlich-Technische Fakult\"at,
Universit\"at Siegen, Siegen, Germany}

@@

Our method is an extension of the classical method of moments with
orthogonal functions~\cite[sec. 8.2]{James:2006zz}. The only difference
is that conventionally the angular functions are assumed
\emph{self-dual}, $\dual{f}_i = f_i$. However, it suffices that the
system of angular functions $\{f_i(\thvec)\}$ \emph{can} be transformed into
an orthonormal basis. We find it convenient to work
in the basis of Legendre polynomials that are not self-dual.
Our approach covers the self-dual case, provided that one
replace $\dual{f}_i \to f_i$ appropriately.
Using the ansatz
\begin{equation}
  \label{eq:dual-ansatz}
  \dual{f}_i = \sum_{j} a_{ij} f_j \,,
\end{equation}
the dual basis needs to be worked out case by case through solving the
linear system of equations (\ref{eq:def-ortho-rel}). For a selection
of hadron decays with a $b$ quark in the initial state and two leptons
in the final state, we list the dual bases in a series of appendices
\ref{app:btokll} through \ref{app:btokstarll}. Note that a similar
analysis was done in~\cite{Dighe:1998vk} for the decays $B \to J/\psi
\phi$ and $B \to J/\psi K^{*}$. In~\cite[section X]{Chung:1997qd}, the
basic ideas and formulae were already given but the latter contained a
number of typographical errors and the presentation was not easy to
understand. Compared to~\cite{Chung:1997qd}, we fill the gaps needed
to apply the method in practice and explain in detail its advantages
and disadvantages on a sound statistical basis.

@@

The orthonormality relations \refeq{def-ortho-rel} imply that a single angular observable $S_i$
can be projected out of the full PDF $P$ as
\begin{equation}
    \label{eq:det-Pi-analytical}
    S_i(\nuvec) = \int_{\Omega} \rmdx{\vec\theta}  P(\nuvec, \thvec) \dual{f}_i(\thvec)\,.
\end{equation}
It is sensible to refer to the angular observable $S_i$ as the
\emph{$f_i$-moment} of the PDF $P$.  We emphasize that a relation of
type \refeq{det-Pi-analytical} holds for any combination of a density
written as in \refeq{def-P} and an orthonormal basis of angular
functions $\lbrace f_i \rbrace$; \ie, there is no unique
basis of angular functions. For the proof we refer to ref. \cite{Dighe:1998vk}.\\

@@

In what follows, we propose a systematic method to unfold all effects
of $\Ekernel$ through MC simulations and the method of moments,
using that $\Ekernel$ can be expanded --- at
least formally --- in Legendre polymials.  For illustration, we
proceed with the explicit example of a uniangular\footnote{%
  The generalization of this section to multiangular PDFs is
  straighforward. It can be achieved by promoting $x$ to a vector,
  promoting the Legendre polynomials to products of independent
  polynomials or spherical harmonics, and promoting the indices
  $i,j,k,m$ to multi-indices.  } PDF with $x = \cos\theta$. Let us
define the PDF in terms of the Legendre polymials (i.e., $f_k(x)
\equiv p_k(x)$) and angular observables $\vec{S}$ as
\begin{equation}
    P\subt(\xt) \equiv P\subt(\xt | \vec{S}) = \sum_k S_k f_k(\xt),
  \end{equation}
where $k = 0, 1, \dots$ denotes an angular-momentum-like index associated with the observables.
Normalization of $P\subt$ is equivalent to choosing $S_0 = 1/2$. Requiring $P\subt(\xt) \ge 0 \,\forall \xt$
implies $|S_k| \leq 1/2$ for $k > 0$. More stringent relations between the
$S_k$ might hold but are of no concern here. For later use, we define
\begin{equation}
  \label{eq:def-Sk}
  S^{(m)}_k \equiv
  \begin{cases}
    1/2 \delta_{k,0}, & m=0\\
    1/2 (\delta_{k,0} + \delta_{k,m}), & m>0 \,,
  \end{cases}
\end{equation}
and note that
\begin{equation}
    P^{(m)}(\xt) \equiv P(\xt | \vec{S}^{(m)})
\end{equation}
is a valid PDF. The dual basis of angular functions follows then from the normalization
and orthogonality of the Legendre polynomials, and one therefore has
\begin{gather}
    \tilde{f}_k(x) = \frac{2 k + 1}{2} f_k(x)\,,\\
    \int_{-1}^{+1} \rmdx{x} \tilde{f}_k(x) f_l(x) = \delta_{k,l}\,.
\end{gather}
We now define the \emph{simulated raw} moments $\vec{Q}^{(m)}$ as
\begin{equation}
    \label{eq:analytic-raw-moments}
    Q_i^{(m)} \equiv \iint \rmdx{\xt} \rmdx{\xd} \tilde{f}_i(\xd) P^{(m)}(\xt) \Ekernel,
\end{equation}
which are instrumental to our recipe. Monte Carlo estimators of these moments can be constructed from specifically crafted detector events $x_\text{d}^{(n,m)}$, $n = 1, \dots, N_\text{d}^{(m)}$ \fred{This should be $N_\text{t}^{(m)}$, right?}, where
\begin{equation}
    x_\text{d}^{(n,m)} \sim P_E^{(m)}(\xd) \equiv \frac{1}{R^{(m)}} \int \rmdx{\xt} P^{(m)}(\xt) \Ekernel\,.
\end{equation}
In words, for each $m$ it is required to generate events from
a toy physical distribution $P^{(m)}$, for which $S_0 = S_m=1/2$ and all other
observables are set to zero. Next, propagate these events through
a detector simulation. The normalization $R^{(m)}$ is chosen such
that $\int \rmdx{\xd} P_E^{(m)}(\xd)  = 1$.  We emphasize that
$R^{(m)}$ can be estimated as $\est{R}^{(m)} = N_\text{d}^{(m)} /
N_\text{t}$, where $N_\text{t}$ corresponds to the number of
simulated true events. \fred{We should clarify: generate $N_\text{t}^{(m)}$ events and $N_\text{d}^{(m)}$ pass the simulation. In principle, you could generate a different number of events $N_\text{t}^{(m)}$ for each $m$.  } The estimators then read
\begin{equation}
    \est{Q}_i^{(m)} \equiv \est{R}^{(m)} \frac{1}{N_\text{d}^{(m)}} \sum_n^{N_\text{d}^{(m)}} \tilde{f}_i(x_\text{d}^{(n,m)})
\end{equation}

@@

\fred{There is a maximum order of acceptance expansion that could have
  an influence. See \cite[below (118)]{Chung:1997qd}. I think it is
  just a consequence of Gaun'ts triple product. This is good news as
  it relieves the experiment to play around and there is less
  arbirariness. I suppose that was Ulrik's question. The only
  arbitrariness comes from cutting on higher partial waves in the
  angular expansion, that's discussed in
  \refsec{systematics:partial-waves}

  We should talk about $E$ not $\eps$ alone. In principle we just
  expand $P$ and $E$ in the polymials and we implicitly estimate the
  ``angular observables'' of the $E$ expansion.}

@@

We thank Ulrik Egede for insightful discussions and helpful comments
on the manuscript. We are grateful to Gudrun Hiller, Martin Jung, and
Konstantinos Petridis for helpful discussions, as well as Claus Grupen
for helpful advice in searching the literature.  Tobias Schlüter made
us aware of existing ideas on our method in~\cite{Chung:1997qd} and
his thesis~\cite[page 123]{Schlueter:2012}. The work of D.v.D has been
supported by the Bundesministerium f\"ur Bildung und Forschung
(BMBF). M.C. has been supported by the ''Diamond Grant'' founded by
the Polish Ministry of Higher Education. F.B. is grateful for the
support by the DFG Cluster of Excellence "Origin and Structure of the
Universe". Part of the numerical simulations have been carried out on
the computing facilities of the Computational Center for Particle and
Astrophysics (C2PAP).
